No meaningful effects of COVID-19 related social media use on well-being

### 2021-12-01 manuscript.tex
- language updates

\abstract{
In times of crisis such as the Corona pandemic citizens need to stay informed about recent events, the latest political decisions, or mandatory protection measures. To this end, many people use various types of media, and increasingly social media. However, because social media are particularly engaging, some find it hard to disconnect and cannot stop `doomscrolling.' In this preregistered study, I investigate whether using social media for COVID-19 related reasons might put personal well-being at risk. To answer this question I analyzed data from the Austrian Corona Panel Project, which consists of 24 waves with overall 3,018 participants. I ran three random effects within between models, controlling for several stable and varying confounders Results showed that the effects of COVID-19 related social media use on well-being were very small, arguably too small to matter. The findings suggest that fears that social media use during times of crisis impairs well-being are likely to be unfounded.
}

@@

During the COVID-19 pandemic,
numerous events unfolded in quick succession.
Several open questions emerged.
How dangerous is the virus?
Is it spreading in my region?
How is it transmitted, and how can I protect myself?
Because for many it was (and still is) a matter of life or death, people aimed to stay informed regarding the latest developments.
Governments around the world implemented safety measures, such as wearing masks, keeping physical distance, or enforcing lockdowns.
In this extraordinary situation, many people used media excessively to attain information, and especially social media were at an all time high (Statista, 2021).

Some people could not stop using social media to learn about COVID-19 related news.
A new phenomenon termed ``doomscrolling'' emerged:
Users were glued to their screens and found it hard to pursue other relevant activities such as working, taking a break, or looking after their children (Klein, 2021).
As doomscrolling increased it became doubtful whether such social media is helpful, or whether it created an additional burden on mental health (Sandstrom, Buchanan, Aknin, \& Lotun, 2021).
These concerns seem justified:
A study with 6,233 people from Germany that was conducted during the pandemic found that ``{[}f{]}requency, duration and diversity of media exposure were positively associated with more symptoms of depression'' (Bendau et al., 2021, p. 283).

As a result, with this study I want to build on this research and investigate whether or not COVID-19 related social media use affected well-being during the pandemic.
To this end, I analyzed a large-scale panel study from the Austrian Corona Panel Project (Kittel et al., 2020).
The panel consists of 24 waves and has an overall sample size of 3018.
The panel study collected a large number of psychological and demographic variables.
I explicitly aimed to investigate the causal effects of COVID-19 related social media use on well-being.

@@

The underlying theories that guided the selection of variables for my analysis are the two-continua model of mental health (Greenspoon \& Saklofske, 2001) and the hierarchical taxonomy of computer-mediated communication (Meier \& Reinecke, 2020).
According to the two-continua model, mental health consists of (a) psychopathology and (b) well-being.
Well-being can be differentiated into subjective and psychological well-being (Diener, Lucas, \& Oishi, 2018).
Whereas subjective well-being emphasizes hedonic aspects such as happiness and joy, psychological well-being addresses eudaimonic aspects such as fulfillment and meaning.
Subjective well-being is primarily about achieving positive affect and avoiding negative affect.
One of the most prominent indicators of well-being is life satisfaction.
In my view, life satisfaction is best thought of as a meta concept that combines psychological and subjective well-being, because it represents a general appraisal of one's life.
Notably, life satisfaction is stable and fluctuates only little, whereas it's the exact opposite for affect (Dienlin \& Johannes, 2020).
To capture well-being in this study I thus build on life satisfaction, positive affect, and negative affect.
Together, this should provide an encompassing perspective on potential media effects.

The hierarchical taxonomy of computer-mediated communication differentiates six levels of how people engage with digital technology.
First, the device (e.g., smartphone); second, the type of application (e.g., social networking site); third, the branded application (e.g., Twitter); fourth, the feature (e.g., status post); fifth, the interaction (e.g., one-to-many); and sixth, the message (e.g., content) (Meier \& Reinecke, 2020).
Whereas the first four levels focus on the \emph{channel}, the last two address the \emph{type} of communication.
To measure social media use for the consumption of COVID-19 related news and topics, I here employ both the channel and the type of communication perspective, which together provides a nuanced understanding of communication.

First, I investigate how different types of communication affect well-being.
Specifically, I differentiate between active and passive use.
I distinguish (a) \emph{reading} (passive), \emph{posting} (active), and \emph{liking and sharing} COVID-19 related posts (both active and passive).
Second, I analyze how using the most prominent branded applications affects well-being, and whether this effect changes across applications.
Branded apps are separate entities with potentially divergent effects.
Twitter might have a different effect as compared to WhatsApp because of their respective affordances.
For example, Waterloo, Baumgartner, Peter, and Valkenburg (2018) found that it's more adequate to express negative emotions on WhatsApp than on Twitter or on Instagram.
The branded applications investigated here are Facebook, Twitter, Instagram, WhatsApp, and YouTube.
Worth noting, this study is not about \emph{general} social media use during times of COVID, but on social media use \emph{focused} on COVID-19 related content.
Examples of such media use include posting thoughts about the pandemic or retweeting COVID-19 related news.

@@

From a theoretical perspective, how could we explain whether COVID-19 related social media use might affect well-being?
In what follows, I outline potential arguments as to why the effect might be positive or negative, direct or indirect, or nonexistent.
In advance, there does not seem to be a clear winner, and it's likely that both positive and negative effects are equally strong.

First, one could assume a \emph{direct} negative effect on well-being, and especially on positive or negative affect, which is more volatile and fluctuating.
Dangers, inequalities, corruption---these were the headlines during the pandemic across many countries worldwide.
If one learns about such events, the initial reaction might be shock, fear, or dismay.
Repeatedly consuming such news can be depressing, perhaps even changing some general perspectives on life.
That said, because not all news was negative, and because many people showed solidarity and compassion, there was also positive and uplifting content, potentially compensating for the negative effects.

@@

On the other hand, one can make the case that overuse might also be beneficial, especially in times of a pandemic---even if the use is mainly COVID-19 related.
Exchanging COVID-19 related messages with friends via WhatsApp might replace the in-person contact one would have otherwise, but which is literally impossible at the time.
In situations where meaningful and functional activities are prohibited, using social media to exchange about COVID-19 related topics might be not the worst idea.
Besides, given that nowadays a large number of experts, scientists, and politicians converse directly on social media, one can get first-hand high quality information on current developments.

Together, the strongest argument to me is that \emph{in general} the effects of social media on well-being are, on average, small at best.
Because this study only looks at \emph{one part} of social media use---namely, COVID-19 related interactions---it is very focused, diminishing the overall potential of the effects even further.
Whether or not using social media for COVID-19 related aspects is detrimental during a pandemic is also not entirely clear.
Therefore, I expect that COVID-19 related communication on social media does not affect well-being in a meaningful or relevant way.

@@

What does this mean practically and how can it be operationalized?
In this study, COVID-19 related social media use was measured on a 5-point scale, ranging from 1 = \emph{never} to 5 = \emph{several times a day}.
Thus, a change of four units in social media use (e.g., a complete stop) should correspond to a noticeable change in well-being.
But what's a noticeable change in well-being?
According to Norman, Sloan, and Wyrwich (2003), people can reliably distinguish \emph{seven} levels of satisfaction with health.
So if satisfaction is measured on a 7-point scale, we would state that a four unit change in social media use should result in a one unit change in life satisfaction.
(For more information, see Method section ``Inference criteria.'')

@@

The hypothesis explicitly states a causal effect.
In non-experimental studies, longitudinal designs help investigate causality.
Using longitudinal designs alone, however, is not sufficient for establishing correct causal statements (Rohrer \& Murayama, 2021).
In addition, we for example also need to control for confounding third variables, and importantly also for \emph{varying} third variables.

@@

Thus, only when controlling for \emph{all} relevant confounders, we can correctly estimate causality without bias (Rohrer, 2018).
Obviously, we can never be entirely sure to have included all confounders, which makes absolute statements regarding causality virtually impossible.
In addition, when determining the overall causal effect, we need to make sure \emph{not} to control for mediating variables (Rohrer, 2018), for doing so would bias our assessment of the causal effect.
Complicating matters further, it is often unclear if a variable is a mediator or a confounder.\footnote{In addition, there also exist colliders, which I don't discuss here and which complicate the issue even further (Rohrer, 2018).}
However, despite all these caveats, when controlling for relevant variables (that aren't mediators), we can be much more certain that we measured causality correctly.
The aim should therefore be to collect as many varying and nonvarying confounders as possible (which it seems is seldom done in our field), while knowing that absolute certainty regarding causality cannot be reached.

When searching for suitable candidates for confounders, we should look out for variables that affect both media use and well-being.
Controlling for these factors isolates the actual effect of social media use on well-being.
We can also control for variables that affect only social media use or well-being.
However, in doing so not much is gained or lost, because the effects of social media use would remain virtually the same (Kline, 2016; but see McElreath, 2021).

In this study, I hence plan to control for the following variables, which either have already been shown to affect both social media use and well-being or which are likely to do so, and which also are not mediators:
gender, age, education, Austria country of birth, Austria country of birth of parents, text-based news consumption, video-based news consumption, residency Vienna, household size, health, living space, access to garden, access to balcony, employment, work hours per week, being in home-office, household income, outdoor activities, satisfaction with democracy, disposition to take risks, and locus of control.
I will not control for variables such as trust in institutions or trust in media, because these variables might be influenced by social media use to a meaningful extent.

Next to including covariates, it is now increasingly understood that causal effects should be analyzed from an internal, within-person perspective (Hamaker, 2014).
If a specific person changes their media diet, we need to measure how this behavior affects \emph{their own} well-being.
Between-person comparisons from cross-sectional data, where participants are interviewed only once, cannot provide such insights.
In this study, I hence differentiate between-person relations from within-person effects.
And as explicated above, to test the hypothesis I thus consider only the within-person effects.

Finally, one precondition of causality is temporal order.
The cause needs to precede the effect.
Finding the right interval between cause and effect is crucial.
For example, if we want to measure the effect of alcohol consumption on driving performance, it makes a big difference if driving performance is measured one minute, one hour, one day, or one week after consumption.
If variables are stable, longer intervals are needed; if they fluctuate, shorter intervals.
In the case of well-being, we need shorter intervals for affect and longer ones for life satisfaction.
Still, choosing the right interval is challenging, because especially short intervals are hard to implement in practice and often require advanced methods such as experience sampling (also known as in situ measurement or ambulant assessment) (Schnauber-Stockmann \& Karnowski, 2020).

In this study, I therefore adopt an intermediate perspective.
I analyze if when a person changes their social media diet, are there simultaneous changes in their well-being?
When additionally controlling for both stable and varying confounders, we can then be more sure that the effect is indeed causal.

@@

The hypotheses, the sample, the measures, the analyses, and the inference criteria (SESOI, p-value) were preregistered on the Open Science Framework.
The (anonymous) preregistration can be accessed here: \url{https://osf.io/87b24/?view_only=b2289b6fec214fa88ee75a18d45c18f3}.
Because in this study I analyzed data from an already existing large-scale data set, all of these steps were done prior to accessing the data.
The preregistration was designed on the basis of the panel documentation online (Kittel et al., 2020).
In some cases I could not execute the analyses as I had originally planned, for example because some properties of the variables only became apparent when inspecting the actual data.
The most relevant deviations are reported below, and a complete list of all changes can be found in the online \href{https://tdienlin.github.io/Austrian_Corona_Panel/index.html}{companion website} (\url{https://tdienlin.github.io/Austrian_Corona_Panel/index.html}).

@@

\hypertarget{inference-criteria}{%
\subsection{Inference criteria}\label{inference-criteria}}

@@

To illustrate what this means in practice, if the 95\% confidence interval falls completely within the null-region (e.g., \emph{b} = .20, {[}95\% CI: .15, .25{]}), the hypothesis that the effect is trivial is supported.
If the effects interval and the null region overlap (e.g., \emph{b} = .30, {[}95\% CI: .25, .35{]}), the hypothesis is not supported and the results are considered inconclusive, while a meaningful negative effect is rejected.
If the confidence falls completely outside of the null-region (e.g., \emph{b} = .40, {[}95\% CI: .35, .45{]}), the hypothesis is rejected and the existence of a meaningful positive effect is supported.

@@

COVID-19 related social media use focused on communication \emph{types} was measured with the three dimensions of (a) reading, (b) liking and sharing, and (c) posting.
The general introductory question was ``How often during the last week have you engaged in the following activities on social media?''
The three items were ``Reading the posts of others with content on the Coronavirus,'' ``When seeing posts on the Coronavirus, I clicked `like,' `share' or `retweet','' ``I myself wrote posts on the Coronavirus on social media.''
Answer options were 1 (\emph{several times per day}), 2 (\emph{daily}), 3 (\emph{several times per week}), 4 (\emph{weekly}), 5 (\emph{never}).
The items were inverted for the analyses.

COVID-19 related social media use focused on \emph{channels} was measured with five variables.
The general introductory question was ``How often in the last week have you followed information related to the Corona-crisis on the following social media?''
The five items were (a) Facebook, (b) Twitter, (c) Instagram, (d) Youtube, and (e) WhatsApp.
Again, the answer options were 1 (\emph{several times per day}), 2 (\emph{daily}), 3 (\emph{several times per week}), 4 (\emph{weekly}), 5 (\emph{never}).
Again, the items were inverted for the analyses.

@@

\begin{figure}
\includegraphics[width=\textwidth]{figures/fig_results_activity} \caption{The effects of various types of social media use on three indicators of well-being. The black estimates show the effects controlled for a large number of covariates (see text; preregistered); the grey estimates are without control variables (exploratory). The SESOI was \textit{b} = |0.30| for life satisfaction and \textit{b} = |0.15| for affect. Hence, all of the reported effects are considered not meaningful.}\label{fig:res-activity}
\end{figure}

Regarding between-person relations, about which no hypotheses were formulated, only three effects didn't include zero.
Respondents who across all waves used social media more frequently than others to read about COVID-19 related posts reported slightly lower levels of positive affect than others (\emph{b} = -0.03 {[}95\% CI \textgreater{} -0.01, -0.06{]}).
Respondents who across all waves used social media more frequently than others to write COVID-19 related posts reported higher levels of negative affect than others (\emph{b} = 0.06 {[}95\% CI 0.09, 0.03{]}).
Interestingly, respondents who across all waves used social media more frequently than others to write COVID-19 related posts also reported slightly higher levels of positive affect than others (\emph{b} = 0.05 {[}95\% CI 0.09, \textless{} 0.01{]}).
However, note that the effect were still completely inside of the null region, hence not large enough to be considered practically relevant.

@@

\begin{figure}
\includegraphics[width=\textwidth]{figures/fig_results_channel} \caption{The effects of using various social media applications on three indicators of well-being. The black estimates show the effects controlled for a large number of covariates (see text); the grey estimates are without control variables. The SESOI was \textit{b} = |0.30| for life satisfaction and \textit{b} = |0.15| for affect. Hence, all of the reported effects are considered not meaningful.}\label{fig:res-channels}
\end{figure}

@@

In what follows, I briefly report some exploratory analyses that weren't preregistered.
First, to contextualize the results reported above, and to see if other variables showed meaningful effects, I also looked at the effect sizes of selected cherry-picked covariates.
Because each variable had different response options, we would need to define a SESOI for each variable, which for reasons of scope I cannot implement here.
Therefore, I report the results of the standardized scales,
which allows for a better comparison across the differently scaled variables.
For what it's worth, as a rough estimate for the SESOI we can build on the typical convention that small effects start at \emph{r} = \textbar.10\textbar.
The results showed that several effects fell outside of the SESOI, were hence considered meaningful.
This includes for example internal locus of control, health, satisfaction with democracy, or exercising.
For an overview, see Figure \ref{fig:res-control}.

\begin{figure}
\includegraphics[width=\textwidth]{figures/fig_results_control_std} \caption{Results of selected covariates. All variables were standardize except 'Male'  and ‘Employed in public service' because there were binary.}\label{fig:res-control}
\end{figure}

@@

In this study I analyzed the effects of COVID-19 related social media use on well-being.
The data come from a panel study with 24 waves and are representative of the Austrian population.
In a random effects model I separated between person relations from within-person effects and controlled for a large number of both stable and varying covariates, aiming to assess causality.
The results showed that within-person effects were trivial.
People who used social media more than usual to learn about COVID-19 didn't show meaningful changes in their well-being.

The results imply that COVID-19 related social media use does not seem to be particularly relevant for well-being.
Other factors among the third variables that were measured revealed larger effects or relations, suggesting that well-being is determined by alternative aspects such as health, satisfaction with democracy, locus of control, or exercising.
According to this study, popular fears that ``doomscrolling'' or overusing social media during times of crises don't seem to be justified.

@@

If anything, two preliminary and subtle trends can be observed.
First, of all the three COVID-19 related social media activities, people who read about the pandemic more than others showed decreased levels of positive affect, and people who actively posted about the pandemic more than others showed slightly increased levels of negative affect.
On the other hand, people who posted more about COVID-19 also showed slightly higher levels of positive affects, so taken together the results are ambivalent.
Second, in terms of media channels, using Twitter more than usual was related to slightly decreased levels of life satisfaction.
Twitter is considered to have more negative affordances and tonality as compared to other networks such as Instagram (Waterloo, Baumgartner, Peter, \& Valkenburg, 2018), which might help explain the results.
Instagram, on the other hand, was related to slightly increased simultaneous levels of life satisfaction.
To speculate, the often-criticized positivity bias on Instagram might have been somewhat beneficial in times of the pandemic.
That said, all these effects were still very small and arguably too small to matter.
But future research might elaborate on these specific relations to probe their stability and relevance.

@@

In this study, COVID-19 related social media use did not meaningfully affect several indicators of well-being, including life satisfaction, positive affect, and negative affect.
However, factors other than social media use were meaningfully related to well-being, such as physical health, exercise, satisfaction with democracy, or believing that one is in control of one's life.
If it's the aim to improve well-being, it might hence be more fruitful not to focus so much on social media but to address other, more pertinent societal problems related to health care, regular exercise, or a functioning democratic system.

>>>>

\abstract{
In times of crisis such as the Corona pandemic citizens need to stay informed about recent events, the latest political decisions, or mandatory protection measures. To this end, many people use various types of media, and increasingly social media. However, because social media are particularly engaging, some find it hard to disconnect and cannot stop `doomscrolling.' In this preregistered study, I investigate whether using social media for COVID-19 related reasons affects psychological well-being. To answer this question I analyzed data from the Austrian Corona Panel Project, which consists of 24 waves with overall 3,018 participants. I ran three random effects within between models, controlling for several stable and varying confounders. Results showed that the effects of COVID-19 related social media use on well-being were very small, arguably too small to matter. The findings suggest that fears that social media use during times of crisis impairs well-being are likely to be unfounded.
}

@@

During the COVID-19 pandemic,
numerous events unfolded in quick succession and several open questions emerged.
How dangerous is the virus?
Is it spreading in my region?
How is it transmitted, and how can I protect myself?
Because for many it was (and still is) a matter of life or death, people aimed to stay informed regarding the latest developments.
Governments around the world implemented safety measures, such as wearing masks, keeping physical distance, or enforcing lockdowns.
In this extraordinary situation, many people used media excessively to attain information, and especially social media were at an all time high (Statista, 2021).

Some people actually couldn't stop using social media to learn about COVID-19 related news.
A new phenomenon emerged, termed ``doomscrolling'':
Users were glued to their screens and found it hard to pursue other relevant activities such as working, taking a break, or looking after their children (Klein, 2021).
As doomscrolling increased it became doubtful whether such social media is helpful, or whether it created an additional burden on mental health (Sandstrom, Buchanan, Aknin, \& Lotun, 2021).
These concerns seem justified:
A study with 6,233 people from Germany conducted during the pandemic found that ``{[}f{]}requency, duration and diversity of media exposure were positively associated with more symptoms of depression'' (Bendau et al., 2021, p. 283).

As a result, with this study I want to build on this research and investigate whether or not COVID-19 related social media use affected well-being during the pandemic.
To this end, I analyzed a large-scale panel study from the Austrian Corona Panel Project (Kittel et al., 2020).
The panel consists of 24 waves and has an overall sample size of 3,018 participants.
The panel study collected a large number of psychological and demographic variables.
I explicitly aimed to investigate the causal effects of COVID-19 related social media use on well-being.

@@

The underlying theories that guided the selection of variables for my analysis are the two-continua model of mental health (Greenspoon \& Saklofske, 2001) and the hierarchical taxonomy of computer-mediated communication (Meier \& Reinecke, 2020).
According to the two-continua model, mental health consists of (a) psychopathology and (b) well-being.
Well-being can be differentiated into subjective and psychological well-being (Diener, Lucas, \& Oishi, 2018).
Whereas subjective well-being emphasizes hedonic aspects such as happiness and joy, psychological well-being addresses eudaimonic aspects such as fulfillment and meaning.
Subjective well-being is primarily about achieving positive affect and avoiding negative affect.
One of the most prominent indicators of well-being is life satisfaction.
In my view, because it represents a general appraisal of one's life, life satisfaction is best thought of as a meta concept that combines psychological and subjective well-being.
Notably, life satisfaction is stable and fluctuates only little, whereas it's the exact opposite for affect (Dienlin \& Johannes, 2020).
To capture well-being in this study I thus build on life satisfaction, positive affect, and negative affect.
Together, this should provide an encompassing perspective on potential media effects.

The hierarchical taxonomy of computer-mediated communication differentiates six levels of how people engage with digital technology.
First, the device (e.g., smartphone); second, the type of application (e.g., social networking site); third, the branded application (e.g., Twitter); fourth, the feature (e.g., status post); fifth, the interaction (e.g., one-to-many); and sixth, the message (e.g., content) (Meier \& Reinecke, 2020).
Whereas the first four levels focus on the communication \emph{channel}, the last two address the communication \emph{type}.
To measure social media use for the consumption of COVID-19 related news and topics, I here employ both the channel and the type of communication perspective, which together provides a nuanced understanding of communication.

First, I investigate how different types of communication affect well-being.
Specifically, I differentiate between active and passive use.
I distinguish (a) \emph{reading} (passive), (b) \emph{posting} (active), and (c) \emph{liking and sharing} COVID-19 related posts (both active and passive).
Second, I analyze how using the most prominent branded applications affects well-being, and whether this effect changes across applications.
Branded apps are separate entities with potentially divergent effects.
Twitter might have a different effect as compared to WhatsApp because of their respective affordances.
For example, Waterloo, Baumgartner, Peter, and Valkenburg (2018) found that it's more adequate to express negative emotions on WhatsApp than on Twitter or on Instagram.
The branded applications investigated here are Facebook, Twitter, Instagram, WhatsApp, and YouTube.
Worth noting, this study is not about \emph{general} social media use during times of COVID, but on social media use \emph{focused} on COVID-19 related content.
Examples of such media use include posting thoughts about the pandemic or retweeting COVID-19 related news.

@@

From a theoretical perspective, how could we explain whether COVID-19 related social media use might affect well-being?
In what follows, I outline potential arguments as to why the effect might be positive or negative, direct or indirect, or nonexistent.
In advance, there doesn't seem to be a clear winner, and it's likely that both positive and negative effects cancel each other out.

First, one could assume a \emph{direct} negative effect on well-being, and especially on positive or negative affect, which are more volatile and fluctuating.
Dangers, inequalities, corruption---these were the headlines during the pandemic across many countries worldwide.
If one learns about such events, the initial reaction might be shock, fear, or dismay.
Repeatedly consuming such news can be depressing, perhaps even changing some general perspectives on life.
That said, because not all news was negative, and because many people showed solidarity and compassion, there was also positive and uplifting content, potentially compensating for the negative effects.

@@

On the other hand, one can make the case that overuse might also be beneficial, especially in times of a pandemic---even if the use is mainly COVID-19 related.
Exchanging COVID-19 related messages with friends via WhatsApp might replace the in-person contact one would have otherwise, but which is literally impossible at the time.
In situations where meaningful and functional activities are prohibited, using social media to exchange about COVID-19 related topics might not be the worst idea.
Besides, given that nowadays a large number of experts, scientists, and politicians converse directly on social media, one can get first-hand high quality information on current developments.

Together, the strongest argument to me is that \emph{in general} the effects of social media on well-being are, on average, small at best.
Because this study only looks at \emph{one part} of social media use---namely, COVID-19 related interactions---it is very focused, diminishing the overall potential of the effects even further.
Whether or not using social media for COVID-19 related aspects is detrimental during a pandemic is also not entirely clear.
Therefore, I expect that COVID-19 related communication on social media doesn't affect well-being in a meaningful or relevant way.

@@

What does this mean practically and how can it be operationalized?
In this study, COVID-19 related social media use was measured on a 5-point scale, ranging from 1 = \emph{never} to 5 = \emph{several times a day}.
Thus, a change of four units in social media use (e.g., a complete stop) should correspond to a noticeable change in well-being.
But what's a noticeable change in well-being?
According to Norman, Sloan, and Wyrwich (2003), people can reliably distinguish \emph{seven} levels of satisfaction with health.
So if satisfaction is measured on a 7-point scale, we would state that a four unit change in social media use should result in a one unit change in life satisfaction.
(For more information, see Methods section ``Inference Criteria.'')

@@

The hypothesis explicitly states a causal effect.
In non-experimental studies, longitudinal designs can help investigate causality.
Using longitudinal designs alone, however, is not sufficient for establishing correct causal statements (Rohrer \& Murayama, 2021).
In addition, we for example also need to control for confounding third variables, and importantly also for \emph{varying} third variables.

@@

Thus, only when controlling for \emph{all} relevant confounders, we can correctly estimate causality without bias (Rohrer, 2018).
Obviously, we can never be entirely sure to have included all confounders, which makes absolute statements regarding causality virtually impossible.
In addition, when determining the overall causal effect, we need to make sure \emph{not} to control for mediating variables (Rohrer, 2018), for doing so would bias our assessment of the causal effect.
Complicating matters further, it is often unclear if a variable is a mediator or a confounder.\footnote{In addition, there also exist colliders, which I don't discuss here and which complicate the issue even further (Rohrer, 2018).}
However, despite all these caveats, when controlling for relevant variables (that aren't mediators), we can be much more certain that we measured causality correctly.
The aim should therefore be to collect as many varying and nonvarying confounders as possible (which I believe is seldom done in our field), while knowing that absolute certainty regarding causality cannot be reached.

When searching for suitable candidates for confounders, we should look for variables that affect both media use and well-being.
Controlling for these factors isolates the actual effect of social media use on well-being.
We can also control for variables that affect only social media use or well-being.
However, in doing so not much is gained or lost, because the effects of social media use would remain virtually the same (Kline, 2016; but see McElreath, 2021).

In this study, I hence plan to control for the following variables, which either have already been shown to affect both social media use and well-being or which are likely to do so, and which also aren't mediators:
gender, age, education, Austria country of birth, Austria country of birth of parents, text-based news consumption, video-based news consumption, residency Vienna, household size, health, living space, access to garden, access to balcony, employment, work hours per week, being in home-office, household income, outdoor activities, satisfaction with democracy, disposition to take risks, and locus of control.
I will not control for variables such as trust in institutions or trust in media, because these variables might be influenced by social media use to a meaningful extent.

Next to including covariates, it's now increasingly understood that causal effects should be analyzed from an internal, within-person perspective (Hamaker, 2014).
If a specific person changes their media diet, we need to measure how this behavior affects \emph{their own} well-being.
Between-person comparisons from cross-sectional data, where participants are interviewed only once, cannot provide such insights.
In this study, I hence differentiate between-person relations from within-person effects.
And as explicated above, to test the hypothesis I thus consider only the within-person effects.

Finally, one precondition of causality is temporal order.
The cause needs to precede the effect.
Finding the right interval between cause and effect is crucial.
For example, if we want to understand the effect of alcohol consumption on driving performance, it makes a big difference if driving performance is measured one minute, one hour, one day, or one week after consumption.
If variables are stable, longer intervals are needed; if they fluctuate, shorter intervals.
In the case of well-being, we need shorter intervals for affect and longer ones for life satisfaction.
Still, choosing the right interval is challenging, because especially short intervals are hard to implement in practice and often require advanced methods such as experience sampling (also known as in situ measurement or ambulant assessment) (Schnauber-Stockmann \& Karnowski, 2020).

In this study, I therefore adopt an intermediate perspective.
I analyze if when a person changes their social media diet, are there \emph{simultaneous} changes in their well-being?
When additionally controlling for both stable and varying confounders, we can then be more sure that the effect is indeed causal.

@@

The hypotheses, the sample, the measures, the analyses, and the inference criteria (SESOI, p-value) were preregistered on the Open Science Framework.
The (anonymous) preregistration can be accessed here: \url{https://osf.io/87b24/?view_only=b2289b6fec214fa88ee75a18d45c18f3}.
Because in this study I analyzed data from an already existing large-scale data set, all of these steps were done prior to accessing the data.
The preregistration was designed on the basis of the panel documentation online (Kittel et al., 2020).
In some cases I couldn't execute the analyses as I had originally planned, for example because some properties of the variables only became apparent when inspecting the actual data.
The most relevant deviations are reported below, and a complete list of all changes can be found in the online \href{https://tdienlin.github.io/Austrian_Corona_Panel/index.html}{companion website} (\url{https://tdienlin.github.io/Austrian_Corona_Panel/index.html}).

@@

\hypertarget{inference-criteria}{%
\subsection{Inference Criteria}\label{inference-criteria}}

@@

Let's briefly illustrate what this means in practice.
If the 95\% confidence interval falls completely within the null-region (e.g., \emph{b} = .20, {[}95\% CI: .15, .25{]}), the hypothesis that the effect is trivial is supported.
If the confidence interval and the null region overlap (e.g., \emph{b} = .30, {[}95\% CI: .25, .35{]}), the hypothesis is not supported and the results are considered inconclusive, while a meaningful negative effect is rejected.
If the confidence interval falls completely outside of the null-region (e.g., \emph{b} = .40, {[}95\% CI: .35, .45{]}), the hypothesis is rejected and the existence of a meaningful positive effect is supported.

@@

COVID-19 related social media use focused on communication types was measured with the three dimensions of (a) reading, (b) liking and sharing, and (c) posting.
The general introductory question was ``How often during the last week have you engaged in the following activities on social media?''
The three items were ``Reading the posts of others with content on the Coronavirus,'' ``When seeing posts on the Coronavirus, I clicked `like,' `share' or `retweet','' ``I myself wrote posts on the Coronavirus on social media.''
Answer options were 1 (\emph{several times per day}), 2 (\emph{daily}), 3 (\emph{several times per week}), 4 (\emph{weekly}), 5 (\emph{never}).
The items were inverted for the analyses.

COVID-19 related social media use focused on channels was measured with five variables.
The general introductory question was ``How often in the last week have you followed information related to the Corona-crisis on the following social media?''
The five items were (a) Facebook, (b) Twitter, (c) Instagram, (d) Youtube, and (e) WhatsApp.
Again, the answer options were 1 (\emph{several times per day}), 2 (\emph{daily}), 3 (\emph{several times per week}), 4 (\emph{weekly}), 5 (\emph{never}).
Again, the items were inverted for the analyses.

@@

\begin{figure}
\includegraphics[width=\textwidth]{figures/fig_results_activity} \caption{The effects of various types of social media use on three indicators of well-being. The black estimates show the effects controlled for a large number of covariates (see text; preregistered); the grey estimates are without control variables (exploratory). The SESOI was b = |0.30| for life satisfaction and b = |0.15| for affect. Hence, all of the reported effects are not considered meaningful.}\label{fig:res-activity}
\end{figure}

Regarding between-person relations, about which no hypotheses were formulated, only three effects did not include zero.
Respondents who across all waves used social media more frequently than others to read about COVID-19 related posts reported slightly lower levels of positive affect than others (\emph{b} = -0.03 {[}95\% CI \textgreater{} -0.01, -0.06{]}).
Respondents who across all waves used social media more frequently than others to write COVID-19 related posts reported higher levels of negative affect than others (\emph{b} = 0.06 {[}95\% CI 0.09, 0.03{]}).
Interestingly, respondents who across all waves used social media more frequently than others to write COVID-19 related posts also reported slightly higher levels of positive affect than others (\emph{b} = 0.05 {[}95\% CI 0.09, \textless{} 0.01{]}).
However, note that the effect were still completely inside of the null region, hence not large enough to be considered practically relevant.

@@

\begin{figure}
\includegraphics[width=\textwidth]{figures/fig_results_channel} \caption{The effects of using various social media applications on three indicators of well-being. The black estimates show the effects controlled for a large number of covariates (see text; preregistered); the grey estimates are without control variables (exploratory). The SESOI was b = |0.30| for life satisfaction and b = |0.15| for affect. Hence, all of the reported effects are not considered meaningful.}\label{fig:res-channels}
\end{figure}

@@

In what follows, I briefly report some exploratory analyses that weren't preregistered.
First, to contextualize the results reported above and to see if the results included any meaningful effects at all, I also looked at the effect sizes of selected (cherry-picked) covariates.
Because each variable had different response options, we would actually need to define a SESOI for each variable, which for reasons of scope I cannot implement here.
Therefore, I report the results of the standardized scales,
which allows for a better comparison across the differently scaled variables.
For what it's worth, as a rough estimate for the SESOI we can build on the typical convention that small effects start at \emph{r} = \textbar.10\textbar.
The results showed that several effects fell outside of the SESOI, were hence considered meaningful.
This includes for example internal locus of control, health, satisfaction with democracy, or exercising.
For an overview, see Figure \ref{fig:res-control}.

\begin{figure}
\includegraphics[width=\textwidth]{figures/fig_results_control_std} \caption{Results of selected covariates. All variables were standardize except 'Male'  and ‘Employed in public service', because there were measured on a binary scale.}\label{fig:res-control}
\end{figure}

@@

In this study I analyzed the effects of COVID-19 related social media use on well-being.
The data come from a panel study with 24 waves and are representative of the Austrian population.
In a random effects model I separated between person relations from within-person effects and controlled for a large number of both stable and varying covariates, thereby aiming to assess causality.
The results showed that within-person effects were trivial.
People who used social media more than usual to learn about COVID-19 didn't show meaningful changes in their well-being.

The results imply that COVID-19 related social media use doesn't seem to be particularly relevant for well-being.
Other factors among the third variables that were measured revealed larger effects or relations, suggesting that well-being is determined by alternative aspects such as health, satisfaction with democracy, locus of control, or exercising.
According to this study, popular fears that ``doomscrolling'' or overusing social media during times of crises don't seem to be justified.

@@

If anything, two preliminary and subtle trends can be observed.
First, of all the three COVID-19 related social media activities, people who read about the pandemic more than others showed slightly decreased levels of positive affect, and people who actively posted about the pandemic more than others showed slightly increased levels of negative affect.
On the other hand, however, people who posted more about COVID-19 also showed slightly higher levels of positive affects, so taken together the results are ambivalent.
Second, in terms of media channels, using Twitter more than usual was related to slightly decreased levels of life satisfaction.
Twitter is considered to have more negative affordances and tonality as compared to other networks such as Instagram (Waterloo, Baumgartner, Peter, \& Valkenburg, 2018), which might help explain the results.
Instagram, on the other hand, was related to slightly increased levels of life satisfaction.
To speculate, the often-criticized positivity bias on Instagram might have been somewhat beneficial in times of the pandemic.
That said, all these effects were still very small and arguably too small to matter.
But future research might elaborate on these specific relations to probe their stability and relevance.

@@

In this study, COVID-19 related social media use didn't meaningfully affect several indicators of well-being, including life satisfaction, positive affect, and negative affect.
However, factors other than social media use were meaningfully related to well-being, such as physical health, exercise, satisfaction with democracy, or believing that one is in control of one's life.
If it's our aim to improve well-being, it might hence be more fruitful not to focus so much on social media but to address other, more pertinent societal problems related to health care, regular exercise, or a functioning democratic system.
